### 2016-06-21 ansi-isolation.tex
Example added in Sec. 3.1.

Having specified isolation levels as trace well-formedness
constraints, we can now construct trace invariants ($\I$) for \txnimp
programs by composing isolation specifications for various
transactions. For instance, the following trace invariant enforces {\sc si}
for both transactions of the program in Fig.~\ref{fig:motiv-eg-1},
allowing it to satisfy its postcondition:
\begin{smathpar}
\I \;=\; \lambda\E.~ \underE{\C{SI(Wd1)}} \conj \underE{\C{SI(Wd2)}}
\end{smathpar}
In contrast, the following trace invariant enforces {\sc rc} for one and {\sc si}
for another, leading to a possible violation of the postcondition:
\begin{smathpar}
\I \;=\; \lambda\E.~ \underE{\C{RC(Wd1)}} \conj \underE{\C{SI(Wd2)}}
\end{smathpar}

>>>>

% Having specified isolation levels as trace well-formedness
% constraints, we can now construct trace invariants ($\I$) for
% \txnimp programs by composing isolation specifications for various
% transactions. For instance, the following trace invariant enforces
% {\sc si} for both transactions of the program in
% Fig.~\ref{fig:motiv-eg-1}, allowing it to satisfy its postcondition:
% \begin{smathpar}
% \I \;=\; \lambda\E.~ \underE{\C{SI(Wd1)}} \conj \underE{\C{SI(Wd2)}}
% \end{smathpar}
% In contrast, the following trace invariant enforces {\sc rc} for one and {\sc si}
% for another, leading to a possible violation of the postcondition:
% \begin{smathpar}
% \I \;=\; \lambda\E.~ \underE{\C{RC(Wd1)}} \conj \underE{\C{SI(Wd2)}}
% \end{smathpar}

### 2016-06-21 motivation.tex
Example added in Sec. 3.1.

\section{Motivation}

@@

For the rely-guarantee approach to be useful, it has to thus constrain
allowed interference in accordance with the chosen level of isolation
associated with a transaction. The first key observation we make is
that we can constrain interference by constraining the definition of
the \emph{happens-before} ($\hbZ$) relation between transactions, or
their respective constituents.  To do so, we axiomatize the $\hbZ$
relation to capture the interference characteristics of different
isolation levels as their specifications ($\psi$). For example, a
\iso{Read Committed} specification allows an operation $\eta_1$ of
$T_1$ to happen before an operation $\eta_2$ of $T_2$ only if every
operation in $T_1$ (including its commit) happens before $\eta_2$ (we
write $T_1 \hboar \eta_2$ to denote this). The {\sc rc} specification
does not require $T_1$ to execute and commit before all operations of
$T_2$, thus allowing it to interfere in $T_2$. The \iso{Snapshot
  Isolation} specification requires two transactions that write to the
same variable to be related by happens-before.  This effectively
prohibits interference due to actions in $T_1$ being interleaved in
$T_2$ (or, vise versa).  Formally\footnote{Note that specifications
  are simplified for illustration. The actual specifications described
  in (\S~\ref{sec:opsem}) are more nuanced.  }:
\begin{smathpar}
\begin{array}{lcl}
\psi_{RC} & \defeq & \forall \eta_1,\eta_2,T_1,T_2.\; \txn(\eta_1) = T_1 
  \conj \txn(\eta_2) = T_2 \\
  & & \hspace*{0.6in}\conj T_1 \neq T_2 \conj \eta_1 \hboar
  \eta_2 \Rightarrow T_1 \hboar \eta_2 \\
\psi_{SI} & \defeq & \forall T_1,T_2.\; T_1 \neq T_2 \conj
  (\exists \C{X}.~{T_1 \wrstoar \C{X}} \conj 
                      {T_2 \wrstoar \C{X}})\\
  &  & \hspace*{0.6in}\Rightarrow{T_1 \hboar T_2} \disj {T_2 \hboar T_1} \\
\end{array}
\end{smathpar}
Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ 
from {\bf Wd2} to {\bf Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. All
arrows are legal under $\psi_{RC}$ because in no case does an operation
$\eta_1$ from {\bf Wd1} happen-before $\eta_2$ of {\bf Wd2} without the commit
of {\bf Wd1} also happening before $\eta_2$. $\psi_{SI}$ however disallows
all dashed arrows, because dashed arrows establish $\hbZ$ between actions in {\bf Wd2}
and only a subset of the operations in {\bf Wd1}. Ligher (blue) arrows with
hollow heads denote $\hbZ$ relationships that do not affect the value
of \C{C} in a way that causes the program to violate its
post-condition.  Darker (black) arrows with solid heads denote $\hbZ$
relationships that lead to the violation of the post-condition. The task of
the reasoning framework is to determine if all $\hbZ$ relationships
allowed by an isolation level lead to the satisfaction of the
post-condition.

>>>>

\section{Motivation}
\label{sec:motivation}

@@

For the rely-guarantee approach to be useful, it has to thus constrain
allowed interference in accordance with the chosen level of isolation
associated with a transaction. The first key observation we make is
that we can constrain interference by constraining the definition of
the \emph{happens-before} ($\hbZ$) relation between transactions, or
their respective constituents.  To do so, we axiomatize the $\hbZ$
relation to capture the interference characteristics of different
isolation levels as their specifications ($\psi$). For example, a
\iso{Read Committed} specification allows an operation $\eta_1$ of
$T_1$ to happen before an operation $\eta_2$ of $T_2$ only if every
operation in $T_1$ (including its commit) happens before $\eta_2$ (we
let $T_1 \hboar \eta_2$ denote this). The {\sc rc} specification
does not require $T_1$ to execute and commit before all operations of
$T_2$, thus allowing it to interfere in $T_2$. The \iso{Snapshot
  Isolation} specification requires two transactions that write to the
same variable to be related by happens-before.  This effectively
prohibits interference due to actions in $T_1$ being interleaved in
$T_2$ (or, vise versa).  Formally\footnote{Note that specifications
    presented here are solely for illustration. The actual
    specifications described in \S\ref{sec:ansi-isolation} are more nuanced.}:
\begin{smathpar}
\begin{array}{lcl}
\psi_{RC} & \defeq & \forall T_1,T_2,\eta_1,\eta_2,.\; \txn(\eta_1) = T_1 
  \conj \txn(\eta_2) = T_2 \\
  & & \hspace*{0.6in}\conj T_1 \neq T_2 \conj \eta_1 \hboar
  \eta_2 \Rightarrow T_1 \hboar \eta_2 \\
\psi_{SI} & \defeq & \forall T_1,T_2.\; T_1 \neq T_2 \conj
  (\exists \C{X}.~{T_1 \wrstoar \C{X}} \conj 
                      {T_2 \wrstoar \C{X}})\\
  &  & \hspace*{0.6in}\Rightarrow{T_1 \hboar T_2} \disj {T_2 \hboar T_1} \\
\end{array}
\end{smathpar}
Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ 
from {\bf Wd2} to {\bf Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. All
arrows are legal under $\psi_{RC}$ because in no case does an operation
$\eta_1$ from {\bf Wd1} happen-before $\eta_2$ of {\bf Wd2} without the commit
of {\bf Wd1} also happening before $\eta_2$. $\psi_{SI}$ however disallows
all dashed arrows, because dashed arrows establish $\hbZ$ between actions in {\bf Wd2}
and only a subset of the operations in {\bf Wd1}. Ligher (blue) arrows with
hollow heads denote $\hbZ$ relationships that do not affect the value
of \C{C} in a way that causes the program to violate its
post-condition.  Darker (black) arrows with solid heads denote $\hbZ$
relationships that lead to the violation of the post-condition. The task of
the reasoning framework is to determine if all $\hbZ$ relationships
allowed by an isolation level lead to the satisfaction of the
post-condition.

### 2016-06-21 opsem.tex
Example added in Sec. 3.1.

We define a small-step operational semantics for this language in
terms of an abstract machine that generates an execution trace
($\E$). The first component of the trace is a set ($\A$) of
\emph{effects}, where an effect ($\eta$) documents a read (\C{RD(X)}),
a write (\C{WR(X)}), or a transaction commit (\C{COMMIT}) operation
performed during the execution. Any value associated with the
operation (\eg, value read or value written) is also documented, and
can be accessed via $\rval$. Every effect has a unique identifier
accessible via $\id$, and a transction identifier accessible via
$\txn$.  The latter identifies the transaction that generated the
effect. In every step of the evaluation, the machine reduces a \txnimp
term by executing a read, write or commit operation, generating an
effect, and extending the trace. Since effects include transaction
identifiers, the semantics distinguishes between terms ($t$) of
different transactions. For example, $\txnbox{t}_i$ denotes a term $t$
inside a transaction $T_i$.  Evaluation contexts are also
appropriately marked. For example, $\ectx_i$ denotes the evaluation
context for a term inside $T_i$. The other component of an execution
trace is a visibility relation ($\visZ$) that establishes a visibility
property between effects among different transactions. The intent and
mechanics of $\visZ$ is described below.

Fundamental to our development is the notion of a trace invariant
($\I$), is a function from traces ($\E$) to propositions
($\texttt{Prop}$) that defines well-formedness constraints over
traces. The machine takes a step only if the resulting trace satisfies
the constraints imposed by $\I$. This behaviour is captured by the
auxiliary reduction rule \rulelabel{E-Aux} that factors out the trace
extension aspect of the evaluation by abstracting away
operation-specific behaviour as a function ($\mathcal{F}$) that
generates an appropriate effect.  \rulelabel{E-Aux} uses $\mathcal{F}$
to generate a new effect and extend the trace ($\E = (\A,\visZ)$)
\emph{only if} the well-formedness constraints imposed by $\I$ on $\E$
(i.e., $\I(\E')$) are satisfied. Otherwise, it gets stuck. In an
execution that runs to completion, every small-step preserves the
well-formedness of a trace, thus ensuring the invariance of $\I$.
Note that the semantics makes no assumptions about $\I$ other than its
type. As such, it can be instantiated with any trace-parametric
proposition that expresses constraints over the given trace. While the
machine takes a step only if the constraints are satisfied, it neither
defines nor explicitly assumes an oracle to check satisfaction.

As described in \S\ref{sec:motivation}, the semantics of various
isolation levels can be captured as constraints over the
happens-before ($\hbZ$) relation. $\hbZ$ is however a derived relation
in our model, composed of more fundamental \emph{session order}
($\soZ$) and \emph{visibility} ($\visZ$) relations.  A session order
relation captures the sequential order of operations within a
transaction. In particular, it relates two effects, $\eta_1$ and
$\eta_2$, such that $\txn(\eta_1) = \txn(\eta_2)$ and $\id(\eta_1) <
\id(\eta_2)$.  The semantics assigns monotonically increasing
identifiers to effects, as defined by the $\id(\eta) > \maxId(\A)$
condition of \rulelabel{E-Aux}.  Evaluation contexts ($\ectx_i$) for
transaction-bound terms are defined so as to enforce a deterministic
sequential order of execution within a transaction, leading to a
deterministic total order among effect ids, which defines the session
order relation. Visibility ($\visZ$) on the other hand relates effects
across concurrent transactions.  Intuitively, $\visZ$ relates $\eta_1$
to $\eta_2$ if and only if $\eta_1$ was \emph{visible} to the
operation that generated $\eta_2$ during its execution, thus effecting
its return value ($\rval(\eta_2)$). For example, a read operation over
\C{X} may pick the value ($\rval$) of the write effect with highest id
among the visible effects (this is made possible by appropriately
defining $\interp{\cdot}$ in \rulelabel{E-Var}, as we show
later). Thus, the value of a read depends on what write effects it can
witness. An operation can only witness the effects of already
concluded operations, which varies between executions due to the
non-deterministic order of evaluating the parallel composition
(cf. evaluation contexts ($\ectx$) for parallel composition) of
transactions. A more notable source of non-determinism, however, is
the \rulelabel{E-Aux} rule, which allows the machine to expose an
arbitrary subset ($S$) of existing effects ($\A$) to the incoming
operation. In other words, the machine is not obligated to reveal the
effects of all previous operations to an incoming operation. This
relaxation allows the abstract machine to model the semantics of
weakly consistent data stores. For instance, operations issued to an
eventually consistent (EC) replicated store could be dispatched to
different replicas whose states may not be in any well-defined
relationship. By allowing operations to witness arbitrary subsets of
the global state, the semantics models the weak visibility properties
of such stores.

However, a machine that lets operations witness arbitrary subset of
the global state offers no isolation whatsoever. For example, it may
allow a read operation to witness writes of an uncommitted
transaction, violating RC isolation. Fortunately, our ability to
express an isolation semantics as constraints over happens-before
order through $\visZ$ and $\soZ$ relations, and the property of the
abstract machine to be parametric over the trace invariant ($\I$),
lets us solve this problem.  In particular, we continue to define
isolation semantics as constraints over $\visZ$ and $\soZ$, but
confine their domain of interpretation to the given trace so that they
now become trace well-formedness constraints. Well-formedness
constraints can be combined into a trace invariant ($\I$). Since the
abstract machine is guaranteed to preserve $\I$ at every step of the
execution, operations are guaranteed to experience the level of
isolation specified by $\I$.  Thus, in executions that run to
completion, the abstract machine models a store that provides the
required levels of isolation. Notably, the machine achieves this
without being restricted to any particular isolation level by relying
on a declarative specification of isolation as trace well-formedness
constraints.  \S\ref{sec:ansi-isolation} shows the specification of
various ANSI SQL isolation levels stated as trace well-formedness
constraints.

>>>>

We define a small-step operational semantics for this language in
terms of an abstract machine that generates an execution trace
($\E$). The first component of the trace is a set ($\A$) of
\emph{effects}, where an effect ($\eta$) documents a read (\C{RD(X)}),
a write (\C{WR(X)}), or a transaction commit (\C{COMMIT}) operation
performed during the execution. Any value associated with the
operation (\eg, value read or value written) is also documented, and
can be accessed via $\rval$. Every effect has a unique identifier
accessible via $\id$, and a transction identifier accessible via
$\txn$.  The latter identifies the transaction that generated the
effect. In every step of the evaluation, the machine reduces a \txnimp
term by executing a read, write or commit operation, generating an
effect, and extending the trace. Since effects include transaction
identifiers, the semantics distinguishes between terms ($t$) of
different transactions. For example, $\txnbox{t}_i$ denotes a term $t$
inside a transaction $T_i$.  Evaluation contexts are also
appropriately marked. For example, $\ectx_i$ denotes the evaluation
context for a term inside $T_i$. The other component of an execution
trace is a visibility relation ($\visZ$) that establishes a visibility
property between effects among different transactions. The intent and
mechanics of $\visZ$ is described in the sequel.

Fundamental to our development is the notion of a trace invariant
($\I$). $\I$ is a function from traces ($\E$) to propositions
($\texttt{Prop}$) that define well-formedness constraints over traces.
The machine takes a step only if the resulting trace satisfies the
constraints imposed by $\I$. This behaviour is captured by the
auxiliary reduction rule \rulelabel{E-Aux} that factors out the trace
extension aspect of the evaluation by abstracting away the
operation-specific behaviour as a function that generates an
appropriate effect. We let $\mathcal{F}$ denote this function.
\rulelabel{E-Aux} uses $\mathcal{F}$ to generate a new effect and
extend the trace ($\E = (\A,\visZ)$) \emph{only if} the
well-formedness constraints imposed by $\I$ on $\E$ (i.e., $\I(\E')$)
are satisfied. Otherwise, it gets stuck. In an execution that runs to
completion, every small-step preserves the well-formedness of a trace,
thus ensuring the invariance of $\I$.  Note that the semantics makes
no assumptions about $\I$ other than its type. As such, it can be
instantiated with any trace-parametric proposition that expresses
constraints over the given trace. For instance, consider the
$\psi_{RC}$ specification from \S\ref{sec:motivation}, but with
bounded $T_1$ and $T_2$ instantiated with \C{Wd1} and \C{Wd2},
respectively. The instantiated specification is the following
proposition:
\begin{smathpar}
\begin{array}{l}
  \forall \eta_1,\eta_2,.\; \txn(\eta_1) = \C{Wd1}
  \conj \txn(\eta_2) = \C{Wd2} \\
  \hspace*{0.6in}\conj \C{Wd1} \neq \C{Wd2} \conj \eta_1 \hboar
  \eta_2 \Rightarrow \C{Wd1} \hboar \eta_2 \\
\end{array}
\end{smathpar}
It is easy to intepret the above proposition in the context of a trace
$\E$ that captures an execution of the program in
Fig.~\ref{fig:motiv-eg-1}. Such a trace-parametric proposition can be
used to instantiate the trace invariant $\I$ in Fig.~\ref{fig:txnimp}.
The resultant operational semantics describes an abstract machine that
gets stuck if an operation of \C{Wd2} is executed in a state that
incorporates some, but not all the effects (including the \C{COMMIT})
of \C{Wd1}. 
% While the machine takes a step only if the constraints are
% satisfied, it neither defines nor explicitly assumes an oracle to
% check satisfaction.

As described in \S\ref{sec:motivation}, the semantics of various
isolation levels can be captured as constraints over the
happens-before ($\hbZ$) relation. $\hbZ$ is however a derived relation
in our model, composed of more fundamental \emph{session order}
($\soZ$) and \emph{visibility} ($\visZ$) relations.  A session order
relation captures the sequential order of operations within a
transaction. In particular, it relates two effects, $\eta_1$ and
$\eta_2$, such that $\txn(\eta_1) = \txn(\eta_2)$ and $\id(\eta_1) <
\id(\eta_2)$.  The semantics assigns monotonically increasing
identifiers to effects, as defined by the $\id(\eta) > \maxId(\A)$
condition of \rulelabel{E-Aux}.  Evaluation contexts ($\ectx_i$) for
transaction-bound terms are defined so as to enforce a deterministic
sequential order of execution within a transaction, leading to a
deterministic total order among effect ids, which defines the session
order relation. Visibility ($\visZ$) on the other hand relates effects
across concurrent transactions.  Intuitively, $\visZ$ relates $\eta_1$
to $\eta_2$ if and only if $\eta_1$ was \emph{visible} to the
operation that generated $\eta_2$ during its execution, thus effecting
its return value ($\rval(\eta_2)$). For example, a read operation over
\C{X} may pick the value ($\rval$) of the write effect with highest id
among the visible effects (this is made possible by appropriately
defining $\interp{\cdot}$ in \rulelabel{E-Var}, as we show
later). Thus, the value of a read depends on what write effects it can
witness. An operation can only witness the effects of already
concluded operations, which varies between executions due to the
non-deterministic order of evaluating the parallel composition of
transactions. A more notable source of non-determinism, however, is
the \rulelabel{E-Aux} rule, which allows the machine to expose an
arbitrary subset ($S$) of existing effects ($\A$) to the incoming
operation. In other words, the machine is not obligated to reveal the
effects of all previous operations to an incoming operation. This
relaxation allows the abstract machine to model the semantics of
weakly consistent data stores. For instance, operations issued to an
eventually consistent (EC) replicated store could be dispatched to
different replicas whose states may not be in any well-defined
relationship. By allowing operations to witness arbitrary subsets of
the global state, the semantics models the weak visibility properties
of such stores.

However, a machine that lets operations witness arbitrary subset of
the global state offers no isolation whatsoever. For example, it may
allow a read operation to witness writes of an uncommitted
transaction, violating RC isolation. Fortunately, our ability to
express an isolation semantics as constraints over happens-before
order through $\visZ$ and $\soZ$ relations, and the property of the
abstract machine to be parametric over the trace invariant ($\I$),
lets us solve this problem.  In particular, we continue to define
isolation semantics as constraints over $\visZ$ and $\soZ$, but
confine their domain of interpretation to the given trace so that they
now become trace well-formedness constraints. Well-formedness
constraints can be combined into a trace invariant ($\I$). Since the
abstract machine is guaranteed to preserve $\I$ at every step of the
execution, operations are guaranteed to experience the level of
isolation specified by $\I$.  Thus, in executions that run to
completion, the abstract machine models a store that provides the
required levels of isolation. Notably, the machine achieves this
without defining an operation semantics for isolation levels, solely
by relying on their declarative characterization as trace
well-formedness constraints.  \S\ref{sec:ansi-isolation} shows the
specification of various ANSI SQL isolation levels stated as trace
well-formedness constraints.
